---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Rmlx

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/hughjonesd/Rmlx/graph/badge.svg)](https://app.codecov.io/gh/hughjonesd/Rmlx)
<!-- badges: end -->

R interface to Apple's MLX (Machine Learning eXchange) library.

## Overview

Rmlx provides an R interface to Apple's [MLX framework](https://ml-explore.github.io/mlx/), enabling high-performance GPU computing on Apple Silicon.  

This package was vibe-coded with Claude/OpenAI Codex in a week.
Use at your own risk! Much of the C++ API has been implemented, but not
python-only features such as large neural network layers.

## Motivation

Modern Macs have a GPU, which is great for performing matrix operations. 
Statistics uses a lot of matrix operations. But until now, there has been no way
for R on the Mac to use the GPU. Rmlx changes that.

## Requirements

- macOS on Apple Silicon *or* Linux with CUDA *or* MacOS/Linux for a CPU-only build

## Installation

`brew install mlx`

or the Linux equivalent. Then just install the package as normal:

```r
# install.packages("remotes")
remotes::install("hughjonesd/Rmlx")
```

Alternatively, you can build mlx from the [source](https://github.com/ml-explore/mlx).

## Features

### Fast GPU Operations

```{r}

library(Rmlx)

A <- matrix(rnorm(1e6), 1e3, 1e3)
system.time(solve(A))
system.time(solve(as_mlx(A)))

```

### Lazy Evaluation

Operations are recorded but not executed until explicitly evaluated:

```{r}

x <- as_mlx(matrix(1:25, 5, 5))
y <- as_mlx(matrix(101:125, 5, 5))

# Lazy - not computed yet
z <- x + y * 2

# Force evaluation
mlx_eval(z)

# Or convert to R (automatically evaluates)
as.matrix(z)

```


### Device Management

M series chips have shared memory between the CPU and GPU, so switching between
devices is costless.

```{r, eval = FALSE}
# Check/set default device
dev <- mlx_default_device()           
mlx_default_device("cpu")    # Switch to CPU
mlx_default_device(dev)      # Back to GPU

# Create on specific device
x_gpu <- as_mlx(matrix(1:12, 3, 4), device = "gpu")
x_cpu <- as_mlx(matrix(1:12, 3, 4), device = "cpu")
```


### Subsetting

Subsetting works like base R:

```{r}
x <- as_mlx(matrix(1:9, 3, 3))
x[1:2, 1:2]

# drop = FALSE by default
x[1, ]

logical_mask <- c(TRUE, FALSE, TRUE)
x[logical_mask, ]

# subset assignment

x[, 2] <- c(0, 0, 0)
x
```

### Arithmetic

```{r}
x <- as_mlx(matrix(1:12, 3, 4))
y <- as_mlx(matrix(13:24, 3, 4))

# Element-wise operations
sum_xy <- x + y
diff_xy <- x - y
prod_xy <- x * y
quot_xy <- x / y
pow_xy <- x ^ 2

# Comparisons
lt <- x < y
eq <- x == y

# Bring results back to R
as.matrix(sum_xy)
as.matrix(lt)
```

### Matrix Operations

Many base R matrix functions have mlx-specific methods:

```{r}
a <- as_mlx(matrix(1:6, 2, 3))
b <- as_mlx(matrix(1:6, 3, 2))

# rbind, cbind, transpose
rbind(a, t(b))
cbind(a, t(b))

# Matrix multiplication
a %*% b

# Reductions
sum(a)
mean(a)
colMeans(a)
rowMeans(a)

# Cumulative operations flatten column-major
as.vector(cumsum(a))

qr_res <- qr(a)
svd_res <- svd(a)
chol_res <- chol(as_mlx(crossprod(matrix(1:6, 3, 2))))
fft_res <- fft(a)

qr_res$Q
svd_res$d
chol_res

```


### Random Sampling

```{r}
mlx_rand_uniform(c(3, 3), min = -1, max = 1)

```

### Data Transformations


```{r}
scores <- as_mlx(c(0.1, 0.7, 0.4, 0.9))
mlx_sort(scores)
mlx_topk(scores, 2)
mlx_argmax(scores)
```

### Matrix Operations


```{r}
A <- as_mlx(matrix(1:4, 2, 2))
B <- as_mlx(matrix(c(1, 0, -1, 0), 2, 2))
A %*% B
solve(A, B)
crossprod(A, B)
```

### Automatic Differentiation

```{r}
loss <- function(w, x, y) {
  preds <- x %*% w
  resids <- preds - y
  sum(resids * resids) / length(y)
}

x <- as_mlx(matrix(rnorm(20), 5, 4))
y <- as_mlx(matrix(rnorm(5), 5, 1))
w <- as_mlx(matrix(0, 4, 1))

grads <- mlx_grad(loss, w, x, y)

# Inspect gradient
as.matrix(grads[[1]])

# Simple SGD loop
model <- mlx_linear(4, 1, bias = FALSE)
opt <- mlx_optimizer_sgd(mlx_parameters(model), lr = 0.1)
loss_fn <- function(mod, data_x, data_y) {
  preds <- mlx_forward(mod, data_x)
  resids <- preds - data_y
  sum(resids * resids) / length(data_y)
}
for (step in 1:50) {
  mlx_train_step(model, loss_fn, opt, x, y)
}

# Check final loss
final_loss <- mlx_forward(model, x)
mean((final_loss - y) * (final_loss - y))
```


## Data Types

Supported data types:

- `float32` for numeric data (default)
- `bool` for logical data
- Integer types `int8`, `int16`, `int32`, `int64`, `uint8`, `uint16`, `uint32`, 
  `uint64`.
- `complex64`





