---
title: "Getting Started with Rmlx"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with Rmlx}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Detect available device
device <- mlx_get_device()
```


## Introduction

Apple MLX (Machine Learning eXchange) is Apple’s high-performance array and machine-learning framework for Apple Silicon, built on top of Metal for GPU execution and optimized CPU kernels. It offers lazy evaluation, vectorized math, automatic differentiation, and neural network building blocks (see the [official MLX documentation](https://ml-explore.github.io/mlx/) for full details).

`Rmlx` is a thin R layer over MLX that lets you:

* Create MLX tensors from R data (`as_mlx()`).
* Run GPU-accelerated math, linear algebra, FFTs, and reductions with familiar R syntax.
* Use automatic differentiation (`mlx_grad()`, `mlx_value_grad()`) for optimization.
* Build simple models with MLX modules and update them using SGD helpers.

All heavy computation stays in MLX land; you only copy back to base R when you call functions like `as.matrix()`.

## System Requirements

Before using Rmlx, ensure MLX is installed:

```bash
# Using Homebrew (if available)
brew install mlx

# Or build from source
git clone https://github.com/ml-explore/mlx.git
cd mlx && mkdir build && cd build
cmake .. && make && sudo make install
```

## Creating MLX Arrays

Convert R objects to MLX arrays using `as_mlx()`:

```{r}
library(Rmlx)

# From a vector
v <- as_mlx(1:10)
print(v)

# From a matrix
m <- matrix(1:12, nrow = 3, ncol = 4)
x <- as_mlx(m)
print(x)

# Specify the device explicitly (uses GPU if available, CPU otherwise)
x_dev <- as_mlx(m, device = device)
```

> **Precision note:** Numeric inputs are stored in single precision (`float32`).
> Requesting `dtype = "float64"` will downcast the input with a warning.
> Logical inputs are stored as MLX `bool` tensors (logical `NA` values are not
> supported). Complex inputs are stored as `complex64` (single-precision real
> and imaginary parts). Use base R arrays if you need double precision arithmetic.

## Lazy Evaluation

MLX arrays use lazy evaluation - operations are recorded but not computed
until needed:

```{r}
# These operations are not computed immediately
x <- as_mlx(matrix(1:100, 10, 10))
y <- as_mlx(matrix(101:200, 10, 10))
z <- x + y * 2

# Force evaluation of a specific array
mlx_eval(z)

# Or convert to R (automatically evaluates)
as.matrix(z)

# Wait for all queued work on the available device
mlx_synchronize(device)
```

## Arithmetic Operations

Rmlx supports standard arithmetic operators:

```{r}
x <- as_mlx(matrix(1:12, 3, 4))
y <- as_mlx(matrix(13:24, 3, 4))

# Element-wise operations
sum_xy <- x + y
diff_xy <- x - y
prod_xy <- x * y
quot_xy <- x / y
pow_xy <- x ^ 2

# Convert back to R to see results
as.matrix(sum_xy)
```

## Matrix Operations

### Matrix Multiplication

```{r}
a <- as_mlx(matrix(1:6, 2, 3))
b <- as_mlx(matrix(1:6, 3, 2))

# Matrix multiplication
c <- a %*% b
as.matrix(c)
```

### Transpose

```{r}
x <- as_mlx(matrix(1:12, 3, 4))
x_t <- t(x)
print(x_t)
```

### Cross Products

```{r}
x <- as_mlx(matrix(rnorm(20), 5, 4))
true_w <- as_mlx(matrix(c(2, -1, 0.5, 0.25), 4, 1))
y <- x %*% true_w
w <- as_mlx(matrix(0, 4, 1))

# Loss must stay entirely in MLX-land: no conversions back to base R
loss <- function(theta, data_x, data_y) {
  preds <- data_x %*% theta
  resids <- preds - data_y
  sum(resids * resids) / length(data_y)
}

grads <- mlx_grad(loss, w, x, y)

# Wrong: converting to base R breaks the gradient
bad_loss <- function(theta, data_x, data_y) {
  preds <- as.matrix(data_x %*% theta)  # leaves MLX
  resids <- preds - as.matrix(data_y)
  sum(resids * resids) / nrow(resids)
}
try(mlx_grad(bad_loss, w, x, y))

# A small SGD loop using the module/optimizer helpers
model <- mlx_linear(4, 1, bias = FALSE)  # learns a single weight vector
parameters <- mlx_parameters(model)
opt <- mlx_optimizer_sgd(parameters, lr = 0.1)
loss_fn <- function(mod, data_x, data_y) {
  theta <- mlx_param_values(parameters)[[1]]
  loss(theta, data_x, data_y)
}

loss_history <- numeric(50)
for (step in seq_along(loss_history)) {
  step_res <- mlx_train_step(model, loss_fn, opt, x, y)
  loss_history[step] <- as.vector(step_res$loss)
}

# Check final loss and inspect learned parameters
final_loss <- mlx_forward(model, x)
residual_mse <- as.vector(mean((final_loss - y) * (final_loss - y)))
residual_mse
loss_history

learned_w <- mlx_param_values(parameters)[[1]]
as.matrix(learned_w)
as.matrix(true_w)
```

## Reductions

Compute summaries across arrays:

```{r}
x <- as_mlx(matrix(1:100, 10, 10))

# Overall reductions
sum(x)
mean(x)

# Column and row means
colMeans(x)
rowMeans(x)

# Convert to R to see values
as.matrix(colMeans(x))

# Cumulative operations flatten the array in column-major order
as.vector(cumsum(x))
```

## Indexing

Subset MLX arrays similar to R:

```{r}
x <- as_mlx(matrix(1:100, 10, 10))

# Select rows and columns
x_sub <- x[1:5, 1:5]

# Select specific row
row_1 <- x[1, ]

# Select specific column
col_1 <- x[, 1]
```

## Device Management

Control whether computations run on GPU or CPU:

```{r}
# Check default device
mlx_default_device()

# Set to CPU for debugging
mlx_default_device("cpu")

# Create array on CPU
x_cpu <- as_mlx(matrix(1:12, 3, 4), device = "cpu")

# Set back to best available device
mlx_default_device(device)
```

Remember that numeric computations are always performed in `float32`; the CPU
mode is useful when you need to compare against base R or debug without a GPU.

## Performance Comparison

Here's a simple timing comparison for large matrix multiplication:

```{r}
n <- 1000

# R base
m1 <- matrix(rnorm(n * n), n, n)
m2 <- matrix(rnorm(n * n), n, n)
t1 <- system.time(r_result <- m1 %*% m2)

# MLX
x1 <- as_mlx(m1)
x2 <- as_mlx(m2)
mlx_eval(x1)
mlx_eval(x2)
t2 <- system.time({
  mlx_result <- x1 %*% x2
  mlx_eval(mlx_result)
  final <- as.matrix(mlx_result)
})

cat("Base R:", t1["elapsed"], "seconds\n")
cat("MLX:", t2["elapsed"], "seconds\n")
```

Note: This is an informal comparison, not a rigorous benchmark. Performance
gains depend on array size, operation type, and hardware.

## Best Practices

1. **Keep data on GPU**: Minimize transfers between R and MLX
2. **Use lazy evaluation**: Build computation graphs before evaluating
3. **Batch operations**: Combine operations before forcing evaluation
4. **Monitor memory**: GPU memory is limited; free unused arrays
5. **Start with CPU**: Use CPU device for debugging, then switch to GPU

## Limitations

Current limitations in this initial version:

- Apple Silicon only (no Intel Mac or other platforms)
- 2D arrays (matrices) are primary focus
- Limited indexing operations
- No autodiff or gradient computation (planned for future release)
