---
title: "Linear Regression with MLX"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Regression with MLX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette demonstrates linear regression using Rmlx, based on the [MLX linear regression example](https://ml-explore.github.io/mlx/build/html/examples/linear_regression.html). We'll train a linear model using automatic differentiation and stochastic gradient descent (SGD) on GPU-accelerated arrays.

## Problem Setup

We'll create synthetic data for linear regression with high dimensionality:
- A random "true" weight vector `w_star` of dimension 100
- A random design matrix `X` of 1000 examples × 100 features
- Noisy labels `y = X @ w_star + small_noise`

```{r}
library(Rmlx)

# Problem metadata
num_features <- 100
num_examples <- 1000
num_iters <- 10000  # iterations of SGD
lr <- 0.01          # learning rate for SGD

# Set seed for reproducibility
set.seed(42)

# True parameters (what we're trying to learn)
w_star <- mlx_rand_normal(c(num_features, 1))

# Input examples (design matrix)
X <- mlx_rand_normal(c(num_examples, num_features))

# Noisy labels
eps <- 1e-2 * mlx_rand_normal(c(num_examples, 1))
y <- X %*% w_star + eps
```



## Define the Loss Function

The mean squared error loss is a standard choice for regression:

```{r}
# Define loss function
loss_fn <- function(w) {
  preds <- X %*% w
  residuals <- preds - y
  0.5 * mean(residuals * residuals)
}
```

The loss measures how well our parameters `w` predict the labels. Lower loss means better predictions.

## Automatic Differentiation

Rmlx provides `mlx_grad()` to compute gradients via automatic differentiation. This computes the gradient of the loss with respect to our parameters:

```{r}
# Get the gradient function
grad_fn <- function(w) {
  mlx_grad(loss_fn, w)[[1]]
}
```

## Training Loop with SGD

We train by repeatedly computing gradients and updating parameters. In each 
iteration, we:

1. Compute the gradient of loss with respect to `w`
2. Update parameters using the gradient step
3. Force evaluation to prevent the computation graph from growing unbounded
4. Monitor progress by printing loss every 1000 iterations


```{r}
# Initialize parameters randomly
w <- 1e-2 * mlx_rand_normal(c(num_features, 1))

# Training loop
for (i in seq_len(num_iters)) {
  # Compute gradient
  grad <- grad_fn(w)

  # Update parameters: w = w - lr * grad
  w <- w - lr * grad

  # Evaluate to avoid building up computation graph
  mlx_eval(w)

  # Print progress
  if (i %% 1000 == 0) {
    loss <- loss_fn(w)
    cat("Iteration", i, "- Loss:", as.vector(loss), "\n")
  }
}
```

## Evaluation

After training, we evaluate how well our learned parameters match the true parameters:

```{r}
# Compute final loss
final_loss <- loss_fn(w)
cat("Final loss:", as.vector(final_loss), "\n")

# Compute error between learned and true parameters
error <- w - w_star
error_norm <- sqrt(sum(error * error))
cat("||w - w*|| =", as.vector(error_norm), "\n")

# Convert to R for detailed inspection
w_learned <- as.matrix(w)
w_true <- as.matrix(w_star)

# Compare first few parameters
print(cbind(learned = w_learned[1:10, 1], true = w_true[1:10, 1]))
plot(w_true[, 1], w_learned[, 1], xlab = "True parameter", 
     ylab = "Learned parameter")
```


## Device Selection

By default, computations run on GPU for speed. Switch to CPU if needed:

```{r}
# Use CPU (useful for debugging)
mlx_default_device("cpu")

# Or back to GPU
mlx_default_device("gpu")
```


